\section{Introduction}

An evolutionary transition in individuality is an event where independently replicating entities unite to replicate as a single, higher-level individual \citep{smith1997major}.
These transitions are understood as essential to natural history's remarkable record of complexification and diversification \citep{smith1997major}.
Likewise, artificial life researchers have highlighted transitions in individuality as a mechanism that is missing in digital systems, but necessary for achieving the evolution of complexity and diversity that we witness in nature \citep{taylor2016open, banzhaf2016defining}.

Fraternal evolutionary transitions in individuality are transitions in which the higher-level replicating entity is derived from the combination of cooperating kin that have entwined their long-term fates \citep{west2015major}.
Multicellular organisms and eusocial insect colonies exemplify this phenomenon \citep{smith1997major} given that both are sustained and propagated through the cooperation of lower-level kin.
This work focuses on fraternal transitions.
Although not our focus here, egalitarian transitions --- events in which non-kin unite, such as the genesis of mitochondria by symbiosis of free-living prokaryotes and eukaryotes \citep{smith1997major} --- also constitute essential episodes in natural history.

In nature, major fraternal transitions occur sporadically with few extant transitional forms, making them challenging to study.
For instance, on the order of 25 independent origins of Eukaryotic multicellularity are known \citep{grosberg2007evolution} with most transitions having occurred hundreds of millions of years ago \citep{libby2014ratcheting}.
Recent work in experimental evolution \citep{ratcliff2014experimental, ratcliff2015origins, gulli2019evolution, koschwanez2013improved}, mechanistic modeling \citep{hanschen2015evolutionary, staps2019emergence}, and digital evolution \citep{goldsby2012task, goldsby2014evolutionary} complements traditional post hoc approaches focused on characterizing the record of natural history.
These systems each instantiate the evolutionary transition process, allowing targeted manipulations to test hypotheses about the requisites, mechanisms, and evolutionary consequences of fraternal transitions.
Digital evolution, computational model systems designed to instantiate evolution in abstract algorithmic substrates rather than directly emulating any specific biological system \citep{wilke2002biology,dolsondigital2022}, occupies a sort of middle ground between wet work and mechanistic modeling.
This approach offers a unique conjunction of experimental capabilities that complements work in both of those disciplines.
Like modeling, digital evolution affords rapid generational turnover, complete observability (every event in a digital system can be tracked), and complete manipulability (every event in a digital system can can be arbitrarily altered).
However, as with \textit{in vivo} experimental evolution, digital evolution systems can exhibit rich evolutionary dynamics stemming from complex, rugged fitness landscapes \citep{labar2017evolution} and sophisticated agent behaviors \citep{grabowski2013case}.

Our work here follows closely in the intellectual vein of Goldsby's deme-based digital evolution experiments \citep{goldsby2012task, goldsby2014evolutionary}.
In her studies, high-level organisms exist as a group of cells within a segregated, fixed-size subspace.
High-level organisms that must compete for a limited number of subspace slots.
Individual cells that comprise an organism are controlled by heritable computer programs that allow them to self-replicate, interact with their environment, and communicate with neighboring cells.

Goldsby's work defines two modes of cellular reproduction: tissue accretion and offspring generation.
In this way, somatic and gametogenic modes of reproduction are explicitly differentiated.
Within a group, cells undergo tissue accretion, whereby a cell copies itself into a neighboring position in its subspace.
In the latter, a population slot is cleared to make space for a daughter organism then seeded with a single daughter cell from the parent organism.

Goldsby's model abstracts away developmental cost to focus on resource competition between groups.
Cells grow freely within an organism, but fecundity depends on the collective profile of computational tasks (usually mathematical functions) performed within the organism.
When an organism accumulates sufficient resource, a randomly chosen subspace is cleared and a single cell from the replicating organism is used as a propagule to seed the new organism.
This setup mirrors the dynamics of biological multicellularity, in which cell proliferation may either grow an existing multicellular body or found a new multicellular organism.

Here, we take several steps to develop a computational environment that removes the enforcement and rigid regulation of multiple organismal levels.
Specifically, we remove the explicitly segregated subspaces and we let multicells interact with each other more freely.
We demonstrate the emergence of multicellularity where each organism manages its own spatial distribution and reproductive process.
This spatially unified approach enables more nuanced interactions among organisms, albeit at the cost of substantially more complicated analyses.
Instead of a single explicit interface to mediate interactions among high-level organisms, such interactions must emerge via many cell-cell interfaces.
Novelty can occur in terms of interactions among competitors, among organism-level kin, or even within the building blocks that make up hierarchical individuality.
Experimentally studying fraternal transitions in a digital system where key processes (reproductive, developmental, homeostatic, and social) occur implicitly within a unified framework can provide unique insights into nature.
For example, pervasive, arbitrary interactions between multicells introduces the possibility for strong influence of biotic selection.

However, in our system, multicells do not emerge from an entirely impartial substrate.
We do explicitly provide some framework to facilitate fraternal transitions in individuality by allowing cells to readily designate distinct hereditary groups.
Offspring cells may either remain part of their parent's hereditary group or found a new group.
Cells can recognize group members, thus allowing targeted communication and resource sharing with kin.
We reward cells for performing tasks designed to require passive collaboration among hereditary group members.
As such, cells that form hereditary groups to maximize advantage on those tasks stand to increase their inclusive fitness.
In previous work introducing the DISHTINY (DIStributed Hierarchical Transitions in IndividualitY) framework we evolved parameters for manually designed cell-level strategies to explore fraternal transitions in individuality \citep{moreno2019toward}.
In this work we extend DISHTINY to incorporate a more dynamic event-driven genetic programming representation called SignalGP, which was designed to facilitate dynamic interactions among agents and between agents and their environment \citep{lalejini2018evolving}.
As expected, with the addition of cell controllers capable of nearly arbitrary computation we see a far more diverse set of behaviors and strategies arise.

Here, we perform case studies to characterize notable multicellular phenotypes that evolved via this more dynamic genetic programming underpinning.
Each case study strain was chosen by screening the entire set of replicate evolutionary runs for signs of the trait under investigation and then manually the most promising strain(s) for further investigation.
Case studies presented therefore represent an anecdotal sampling, rather than an exhaustive summary, with respect to each trait of interest.
Our goal is to explore a breadth of possible evolutionary outcomes under the DISHTINY framework.
We see this as a precursory step toward hypothesis-driven work contributing to open questions about fraternal transitions in individuality.
