\section{Materials and Methods}

\input{fig/signalgp-dishtinygp.tex}

We performed simulations in which cells evolved open-ended behaviors to make decisions about resource sharing, reproductive timing, and apoptosis.
We will first describe the environment and hereditary grouping system cells evolved under and then describe the behavior-control system cells used.

\subsection{Cells and Hereditary Groups}

Cells occupy individual tiles on a 60-by-60 toroidal grid.
Over discrete time steps (``updates''), cells can collect a resource.
Collected resource decays at a rate of 0.1\% per update, incentivizing its quick use but gradual enough so as to not prevent the most naive cells from eventually accumulating enough resource to reproduce.
Once sufficient resource accrues, cells may pay one unit of resource to place a daughter cell on an adjoining tile of the toroidal grid (i.e., reproduce), replacing any existing cell already there.
Daughter cells inherit their parent's genetic program, except any novel mutations that may arise.
We used standard SignalGP mutation parameters from \citep{lalejini2018evolving}, but only applied mutations 1\% of daughter cells at birth.
Daughter cells may also inherit hereditary group ID, introduced and discussed below.

Cells accrue resource via a cooperative resource-collection process.
The simulation distributes large amounts of resource within certain spatial bounds in discrete, intermittent events.
Working in a group allows cells to more fully collect available resource during these events.
Cooperating in medium-sized groups (on the order of 100 cells) accelerates per-cell resource collection rate.
Unicellular, too-small, or too-large groups collect resource at a lesser per-cell rate.
As an arbitrary side effect of the simulation algorithm employed to instantiate the cooperative resource distribution process, groups with a roughly circular layout collect resource faster than irregularly-shaped groups.
Cooperative resource collection unfolds as an entirely passive process on the part of the cells, influenced only by a group's spatial layout.
Full details on the simulation algorithm that determines cooperative resource collection rates appear in in Supplementary Section \ref{sup:resource_collection_process}.

Cells may grow a cooperative resource-collecting group through cell proliferation.
We refer to these cooperative, resource-collecting groups as ``hereditary groups.''
As cells reproduce, they can choose to adsorb daughter cells onto the parent's hereditary group or expel those offspring to found a new hereditary group.
These decisions affect the spatial layout of these hereditary groups and, in turn, affect individual cells' resource-collection rate.

To promote group turnover, we counteract the advantage of established hereditary groups with a simple aging scheme.
As hereditary groups age over elapsed updates and somatic generations, their constituent cells lose the ability to regenerate somatic tissue and then, soon after, to collect resource.
A complete description of group aging mechanisms used appears in Supplementary Section \ref{sup:hereditary_group_life_cycle}.

Because new hereditary group IDs arise first in a single cell and grow disseminate exclusively among direct descendants of that progenitor cell, hereditary groups are reproductively bottlenecked.
This clonal (or ``staying together'') multicellular life history stands in contrast with an aggregative (or ``coming together'') life cycle where chimeric groups arise via fusion of potentially loosely-related lineages \citep{staps2019emergence}.
Such clonal development is known to strengthen between-organism selection effects \citep{grosberg2007evolution}.

In this work, we screen for fraternal transitions in individuality with respect to these hereditary groups by evaluating three characteristic traits of higher-level organisms: resource sharing, reproductive division of labor, and apoptosis.
We can further screen for the evolution of complex multicellularity by assessing cell-cell messaging, regulatory patterning, and functional differentiation between cells within hereditary groups \cite{knoll2011multiple}.

\subsection{Hierarchical Nesting of Hereditary Groups} \label{sec:hierarchical_nesting}

Successive fraternal transitions in natural history --- for example, to multicellularity and then to eusociality \citep{smith1997major} --- underscores the constructive power of evolution to harness emergent structures as building blocks for further novelty.
To explore these dynamics, in some experimental conditions we incorporated a hierarchical extension to the hereditary grouping scheme described above.

Hierarchical levels are introduced into the system by providing a mechanism to groups of hereditary groups to form.
We accomplish this through two separate, but overlaid, instantiations of the hereditary grouping scheme.
We refer to each independent hereditary grouping system as a ``level.''
The hierarchical extension allows two levels of hereditary grouping, identified here as L0 and L1.
L0 instantiates smaller, inner grouping embedded inside of a L1 grouping.
Without the hierarchical extension, only L0 is present.
\footnote{
We chose to number these levels using the computer science convention of zero-based indexing (as opposed to everyday practice of counting up from one) to maintain consistency with source code and data sets associated with this work.
}
We refer to the highest hereditary grouping level present in a simulation as the ``apex'' level.

Under the hierarchical extension, each cell contained a pair of separate hereditary group IDs --- the first for L0 and the second for L1.
During reproduction, daughter cells could either
\begin{enumerate}
\item inherit both L0 and L1 hereditary group ID,
\item inherit L0 hereditary group ID but not L1 hereditary group ID, or
\item inherit neither hereditary group ID.
\end{enumerate}
In order to enforce hierarchical nesting of hereditary group IDs, daughter cells could not inherit just the L1 hereditary group ID.

Hierarchically hereditary group IDs are strictly nested: all cells are members of one L0 hereditary group and L1 hereditary group.
No cell can be a member of two L0 hereditary groups or two L1 hereditary groups.
Likewise, no L0 hereditary groups can appear within two different L1 hereditary groups.
Useful as a concrete illustration of this scheme, Figure \ref{fig:ko-morphology}\textbf{(A)} depicts hierarchically-nested hereditary groupings assumed by an evolved strain.

\subsection{Cell-Level Organisms}

Our experiments use cell-level digital organisms controlled by genetic programs subject to mutations and selective pressures that stem from local competition for limited space.

We employ the SignalGP event-driven genetic programming representation.
As sketched in Figure \ref{fig:signalgp-dishtinygp}\textbf{(A)}, this representation is specially designed to express function-like modules of code in response to internal signals or external stimuli.
This process can be considered somewhat akin to gene expression.
In our experiments, virtual CPUs can execute responses to up to 24 signals at once, with any further signals usurping the longest-running signal handlers.
The event-driven framework facilitates the evolution of dynamic interactions between digital organisms and their environment (including other organisms) \citep{lalejini2018evolving}.

Special module components allow evolving programs to sense and interact with their environment, through mechanisms including resource sharing, hereditary group sensing, apoptosis, cell reproduction, and arbitrary cell-cell messaging.
Modules can also include general purpose computational elements like conditionals and loops, which allows cells to evolve sophisticated behaviors conditioned on current (and even previous) local conditions.
A simple ``regulatory'' system provides special CPU instructions that dynamically adjust which modules are activated by particular signals.
In our simulation, directionality of some inputs and outputs must be accounted for (e.g., specifying \textit{which} neighbor to share resource with).
To accomplish this, we provide each cell an independent SignalGP hardware instance to manage inputs and outputs with respect to each specific cell neighbor.
So there are four hardware sets per cell, one for each cardinal direction.
\footnote{
This approach differs from previous work evolving digital organisms in grid-based problem domains \citep{goldsby2014evolutionary, goldsby2018serendipitous, grabowski2010early, biswas2014causes, lalejini2018evolving}; see Supplemental Section \ref{sup:cell_level_organisms} for further detail.
}
Figure \ref{fig:signalgp-dishtinygp}\textbf{(B)} overviews the configuration of the four SignalGP instances that constitute a single cell.

Supplementary Sections \ref{sup:cell_level_organisms}, \ref{sup:standard_instruction_library}, \ref{sup:custom_instruction_library}, and \ref{sup:environmental_cue_library} provide full details of the digital evolution substrate underpinning this work.

\subsection{Surveyed Evolutionary Conditions}

To broaden our exploration of possible evolved multicellular behaviors in this system, we surveyed several evolutionary conditions.

In one manipulation, we explored the effect of enabling hierarchical structure within hereditary groups, such that parent cells can choose to keep offspring in their same sub-group, in just the same full group, or expel them entirely to start a new group.
Cells can sense and react to the level of hereditary ID commonality shared with each neighbor.
This manipulation presents opportunity for hierarchical individuality or for a mechanism to mediate differentiation within a multicell, but does not enforce it.

In a second manipulation, we explored the importance of explicitly selecting for medium-sized groups (as had been needed to maximize resource collection) by removing this incentive.
Instead, the system distributed resource at a uniform per-cell rate.

We combined these two manipulations to yield four surveyed conditions:
\begin{enumerate}
\item ``Flat-Even'': One hereditary group level (flat) with uniform resource inflow (even). In-browser simulation: \url{https://hopth.ru/i},
\item ``Flat-Wave'': One  hereditary group level (flat) with group-mediated resource collection (wave); In-browser simulation: \url{https://hopth.ru/j}),
\item ``Nested-Even'': Two hierarchically-nested hereditary group levels (nested) with uniform resource inflow (even). In-browser simulation: \url{https://hopth.ru/k},
\item ``Nested-Wave'': Two hierarchically-nested hereditary group levels (nested) with group-mediated resource collection (wave). In-browser simulation: \url{https://hopth.ru/l}.
\end{enumerate}

Supplementary Section \ref{sup:treatments} provides full details for each of the four surveyed evolutionary conditions.

For each condition, we simulated 40 replicate populations for up to
1,048,576 ($2^{20}$) updates.
During this time, on the order of 4,000 cellular generations and 500 apex-level group generations elapsed in runs.
(Full details appear in Supplementary Table \ref{sup:systematics}.)
Due to variability in simulation speed, four replicates only completed 262,144 updates.
All analyses involving inter-replicate comparisons were therefore performed at this earlier time point.
